Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-05-17 doc/mc-paper.tex
Partial revisions

We present \emph{Partitioned-Operation Graphs} (POGs), a generalization
of the representations used by existing knowledge compilers.
We have designed  CPOG, a framework that can express proofs of equivalence between a
POG  and a Boolean formula in conjunctive normal form (CNF).

We have developed a program that generates POG representations from
the d-DNNF graphs produced by the state-of-the-art knowledge compiler
\dfour{}, as well as checkable CPOG proofs certifying that the output POGs
are equivalent to the input CNF formulas.  Our toolchain
for generating and verifying POGs scales to all but the largest d-DNNF
graphs produced by \dfour{} for formulas from a recent model counting
competition. Additionally, we have developed a formally verified CPOG
checker and model counter for POGs in the \lean{} proof assistant.
In doing so, we proved the soundness of our proof framework. These programs
comprise the first formally verified toolchain for weighted and unweighted
model counting.
\end{abstract}

@@

One approach to model counting, known as {\em knowledge compilation},
transforms the formula into a structured form for which model counting
is straightforward.  For example, the {\em deterministic,
  decomposable, negation normal form} (d-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by a Boolean And or Or operation.  Restrictions
are placed on the structure of the graph such that a count of the
models can be computed by a single bottom-up traversal.
Kimmig et al.\ present a very general {\em
  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
properties of Boolean functions that can be efficiently computed from
a d-DNNF representation.  These include standard and weighted model
counting, and much more.

@@

This paper introduces \emph{Partitioned-Operation Graphs} (POGs),
a generalization of the representations produced by current knowledge
compilers. The CPOG (for ``certified'' POG) file format then
captures both the structure of a POG
and a checkable proof of its logical equivalence to a Boolean formula in
conjunctive normal form (CNF).  A CPOG
proof consists of a sequence of clause addition and deletion steps,
based on an extended resolution proof system~\cite{Tseitin:1983}.
We establish a set of conditions that, when satisified by a CPOG file, guarantee that it
encodes a well-formed POG and provides a valid equivalence proof.

@@

Figure~\ref{fig:chain} illustrates our certifying knowledge
compilation and model counting toolchain.  Starting with
input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
generates a d-DNNF representation, and the {\em proof
  generator} uses this to generate a CPOG file.
The {\em proof checker} verifies the equivalence of the CNF and CPOG representations.
The {\em ring evaluator} computes
a standard or weighted model count from the POG\@ representation.
As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
moves the root of trust away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will be caught by the proof checker.

@@

We evaluate our toolchain using benchmark
formulas from the 2022 standard and weighted model competitions.  Our
tools handle all but the largest d-DNNF graphs generated
by \dfour{}.  We also measure the benefits of several optimizations as
well as the relative performance of the verified checker with one
designed for high performance and capacity.

@@

Generating proofs of unsatisfiability in SAT solvers has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently checked
\cite{RAT,wetzler14_drattrim}.
A number of formally verified checkers have been developed within different verification frameworks~\cite{cruz-cade-2017,lrat,Lammich:20,Tan:2021}.
The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
By contrast, our proofs construct a new propositional formula, and we must verify
%both
that it
%satisfies a set of rules and that it
is equivalent to the input formula.
This requires verifying additional proof steps, including clause deletion steps, and subtle invariants, as described in
Sections~\ref{sect:cpog} and \ref{section:formally-verified-toolchain}.

Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithms commonly used by model counters.
They present a series of lemmas and a theorem stating
that their checking framework is sound.  They also present
experimental results for cases where they modify existing model
counters to generate proof traces, and where they generate proof
traces from d-DNNF graphs.
%% They have not formally verified their
%%proof framework or created a formally verified checker.
Our toolchain is not directly comparable to theirs---they directly
certify the numeric value produced by a model counter, while we
certify that the graph generated by a knowledge compiler is logically
equivalent to the input formula.  We can then validate a standard or
weighted count from that representation.
%% Their tool
%% chain can be applied to some programs that are pure model counters,
%%while ours can be applied to pure knowledge compilers.
Nonetheless,
we compare the performance of our toolchain to theirs as part of our
experimental evaluation in Appendix~\ref{app:experiments}.

We know of no other
work on certifying the output of a knowledge compiler.

@@

POGs generalize the d-DNNF graphs in two ways:
\begin{itemize}
\item They allow negation of arbitrary nodes in the graph, not just
  variables.
  %, and so we allow it in the interest of generality.
%%  Negation could be useful in some contexts, such as when
%%  converting binary decision diagrams (BDDs) having complemented
%%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.
\item They allow arbitrary arguments to a $\por$ operation, as long as
  these have disjoint models.  By contrast, each
  sum node $\nodeu$ in a d-DNNF must have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
%%Our generalization is provided to allow an
%%  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
%%  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  These generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

@@

%% The proof system is based on extended
%% resolution~\cite{Tseitin:1983} with reverse unit propagation
%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
%% proving that a set of clauses logically implies another clause.
%% Alternate version
%% knowledge compilation.  The proof system is based on extended
%% resolution~\cite{Tseitin:1983} with reverse unit propagation
%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
%% proving that a set of clauses logically implies another clause.
%%
%% As an example, consider the formula $\varphi = (\obar{a} \lor b) \land (\obar{b} \lor c \lor \obar{e}) \land (\obar c \lor d)$
%% and a RUP step to derive the target clause $(\obar{a} \lor d \lor \obar{e})$ from the three clauses from $\varphi$.
%% A RUP proof would take the following form.  We start with the assignment that falsifies the target clause. This assignment
%% is extended by clauses in $\varphi$ that become unit and eventually the empty clause. In the illustration below, the
%% involved clauses (antecedents) are shown in the order that they became unit / falsified.
%%
%% \begin{center}
%%   \begin{tabular}{lcccc}
%%          & \makebox[15mm]{Target}    & \makebox[10mm]{} & \makebox[10mm]{Antecedents} & \makebox[10mm]{} \\
%%   Clause & $\obar{a} \lor d \lor \obar{e}$ & $\obar{c} \lor d$  & $\obar{e} \lor \obar{b} \lor c$   & $\obar{a} \lor b$ \\
%%   \midrule
%%   Units  & $a$, $\obar{d}$, $e$      &  $\obar{c}$             & $\obar{b}$                         & $\nil$ \\
%%   \end{tabular}
%% \end{center}
%%
%% RUP is an alternative formulation of resolution.  For target clause
%% $C$, it can be seen that applying resolution operations to the
%% antecedent clauses from right to left will derive a clause $C'$ such
%% that $C' \subseteq C$.  By {\em subsumption} %~\cite{philipp:lpar:2017},
%% we then have $C' \rightarrow C$.  Compared to listing each resolution
%% operation as a separate step, using RUP as the basic proof step makes
%% the proofs more compact.
%%
%% As is shown in Figure~\ref{fig:chain}
%% two programs are involved in producing a certified compilation of a formula:
%% \begin{itemize}
%% \item The {\em proof generator} produces a CPOG file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
%% As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a d-DNNF representation of the formula.
%% \item The {\em proof checker} ensures that all of the proof conditions are satisfied.
%% \end{itemize}
%% >>>>>>> 9c1793da858d00e5d4022d6d33c009848ba2083b

@@

The sequence of operator declarations, asserted clauses, and
clause deletions represents a systematic transformation of the input formula
into a POG formula.  Validating all of these steps serves to prove that the
POG is logically equivalent to the input formula.
At the completion of the proof, the following \textsc{final conditions} must hold:
\begin{enumerate}
\item There is exactly one remaining clause that was added via RUP
  addition, and this is a unit clause consisting of root literal $r$.
\item All of the input clauses have been deleted.
\end{enumerate}
In other words, at the end of the proof it must hold that the active clauses be exactly those
in $\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting
of unit clause $\{r\}$ and the defining clauses of the POG nodes. Recognizing that
any total assignment to the input variables implicitly defines the assignments to the extension variables,
we can see that $\pogformula$ is the clausal encoding of $\phi_\noder$.
Let $\inputformula$ denote the input formula.
The sequence of clause addition steps provides a {\em forward implication} proof that
$\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$.  That is, any total
assignment $\assign$ satisfying the input formula must also satisfy
the formula represented by the POG\@.
Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must
falsify $\phi_\noder$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
that
$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.

@@

\section{Generating CPOG from d-DNNF}
\label{section:generating:cpog}

%%Figure~\ref{fig:chain} illustrates one method of generating a CPOG
%%file from an input formula in conjunctive normal form.  Our toolchain
%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
%%but it would also work for other tools that use a top-down search
%%strategy to generate a d-DNNF representation.
A d-DNNF
graph can be directly translated  into a POG,
although in doing this conversion,
our program also performs simplifications to
eliminate Boolean constants.
Except in degenerate cases,
%where the formula is unsatisfiable or a tautology,
we can therefore assume
that the POG does not contain any constant nodes.
In addition, negation is only
applied to variables, and so we can merge the negation operations into the terminal nodes,
viewing the POG as consisting
of {\em literal} nodes corresponding to input variables and their negations, along with
{\em nonterminal} nodes, which can be further classified as {\em product} and {\em sum} nodes.

@@

The process for generating such a proof depends on the form of node $\nodeu$:
\begin{enumerate}
\item If $\nodeu$ is a literal $\lit'$, then the formula
  $\simplify{\psi}{\passign}$ must consist of the single unit clause
  $C = \{\lit'\}$, such that any $C' \in \prov_{\passign}(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
  Any of these can
  serve as the target clause.
\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
  then, since the node originated from a d-DNNF graph, there must be
  some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
  product node containing a literal node for $x$ as a child.  In either case, we
  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
  This will cause the addition of the target clause
  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
Similarly, either $\nodeu_0$ is a literal node for $\obar{x}$ or $\nodeu_0$ is a product node containing a literal node for $\obar{x}$ as
  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
  causing the addition of the target clause
  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
  These recursive results can be combined with the second and third defining clauses for $\nodeu$
(see Table~\ref{tab:defining}(B))
  to generate the target clause for $\nodeu$.
\item If $\nodeu$ is a product node, then we can divide its children
  into a set of literal nodes $\lambda$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
  \begin{enumerate}
    \item For each literal
  $\lit \in \lambda$, we must prove that any total assignment $\alpha$, such that
  $\passign \subset \alpha$ has $\alpha(\lit) = 1$.  In some
  cases, this can be done by simple Boolean constraint propagation (BCP).
  In other cases, we must prove that the formula
  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
  do so by writing the formula to a file, invoking a proof-generating
  SAT solver, and then converting the generated unsatisfiability proof
  into a sequence of clause additions in the CPOG file.
\item For a single nonterminal child ($k = 1$), we recursively call
  $\validate \left(\nodeu_1, \passign \cup \lambda, \psi\right)$.
\item For multiple nonterminal children ($k > 1$),
  it must be the case that the clauses in
  $\gamma = \simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
  literals.
  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov_{\passign}(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
  We recursively call
  $\validate \left(\nodeu_i, \passign \cup \lambda, \psi_i\right)$.
\end{enumerate}
  We then generate the target clause for node $\nodeu$,
creating the hint by combining the results from the BCP and SAT calls for
  the literals, the recursively computed target clauses, and all but
  the first defining clause for node $\nodeu$
(see Table~\ref{tab:defining}(A)).
\end{enumerate}
Most of these steps involve a polynomial number of
operations per recursive call, with the exception of those that call
a SAT solver to validate a literal.

@@

Each input clause deletion can be done as a single RUP step.  The
proof generator constructs the hint sequence from the defining
clauses of the POG nodes via a single, bottom-up pass through the
graph.  The RUP deletion proof for input clause $C$ effectively proves that any
total assignment $\assign$ that satisfies the POG definition clauses
but does not satisfy $C$ will yield
$\assign(r) = 0$.  It starts with the set of literals
$\{ \obar{\lit} \mid \lit \in C\}$, describing the required condition for
assignment $\assign$ to falsify clause $C$.
It then
adds literals via unit propagation until a
conflict arises.    Unit literal $r$ gets
added right away, setting up a potential conflict.

@@

We summarize the results of our experiments here.  
Appendix~\ref{app:experiments} provides a more complete description. For our
evaluation, we used the public benchmark problems from the 2022
standard and weighted model competitions.  We found that there were
180 unique CNF files among these, ranging in size from 250 to
2,753,207 clauses.
%%We ran our programs on a processor with 64~GB of
%%memory and having an attached high-speed, solid-state
%%disk.
With a runtime limit of 4000 seconds, \dfour{} completed for 124 of the
benchmark problems.  Our proof generator was able to convert all of
these into POGs, with their declarations ranging from 304 to
2,761,457,765 (median 848,784) defining clauses.
%(The
%maximum count overflowed the overflowed the 32-bit signed integers we
%used to represent clause identifiers in both the proof generator and
%checker.)
We ran our proof generator with a time limit of 10,000 seconds.
It was able to generate full proofs for 108 of the problems and
one-sided proofs for an additional 9 of them, leaving just 7 with no
verification.  The prototype checker successfully verified all of the generated
proofs.  The longest runtime for the combination of proof generator
and checker for a full proof was 13,145 seconds.
It is worth noting that, to date, we have not found
any errors in the d-DNNF graphs generated by \dfour{}.

@@

We are hopeful that having checkable proofs for knowledge compilers
will allow them to be used in applications where high levels of trust
are required, and that it will provide a useful tool for developers of
knowledge compilers.
%%Although our current implementation only handles
%%the outputs of one particular program, it could be adapted
Our experiments demonstrate that our toolchain can already
 handle problems nearly at the limits of current knowledge compilers.
Further engineering and optimization of our proof
generator and checker could improve their performance and capacity substantially.

>>>>

We present \emph{Partitioned-Operation Graphs} (POGs), a general form that can
encode all
of the representations used by existing knowledge compilers.
We have designed  CPOG, a framework that can express proofs of equivalence between a
POG  and a Boolean formula in conjunctive normal form (CNF).

We have developed a program that generates POG representations from
the decision-DNNF graphs (a restricted form of d-DNNF)
produced by the state-of-the-art knowledge compiler
\dfour{}, as well as checkable CPOG proofs certifying that the output POGs
are equivalent to the input CNF formulas.  Our toolchain
for generating and verifying POGs scales to all but the largest
graphs produced by \dfour{} for formulas from a recent model counting
competition. Additionally, we have developed a formally verified CPOG
checker and model counter for POGs in the \lean{} proof assistant.
In doing so, we proved the soundness of our proof framework. These programs
comprise the first formally verified toolchain for weighted and unweighted
model counting.
\end{abstract}

@@

One approach to model counting, known as {\em knowledge compilation},
transforms the formula into a structured form for which model counting
is straightforward.  For example,the {\em deterministic
decomposable, negation normal form} (d-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} (as well as the more restrictive {\em decision-DNNF} representation~\cite{oztok:cp:2014})
 represents a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by a Boolean And or Or operation.  Restrictions
are placed on the structure of the graph such that a count of the
models can be computed by a single bottom-up traversal.
Kimmig et al.\ present a very general {\em
  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
properties of Boolean functions that can be efficiently computed from
a d-DNNF representation.  These include standard and weighted model
counting, and much more.

@@

This paper introduces \emph{Partitioned-Operation Graphs} (POGs),
a general form that can encode all of the representations produced by current knowledge
compilers. The CPOG (for ``certified'' POG) file format then
captures both the structure of a POG
and a checkable proof of its logical equivalence to a Boolean formula in
conjunctive normal form (CNF).  A CPOG
proof consists of a sequence of clause addition and deletion steps,
based on an extended resolution proof system~\cite{Tseitin:1983}.
We establish a set of conditions that, when satisified by a CPOG file, guarantee that it
encodes a well-formed POG and provides a valid equivalence proof.

@@

Figure~\ref{fig:chain} illustrates our certifying knowledge
compilation and model counting toolchain.  Starting with
input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
generates a decision-DNNF representation,
and the {\em proof
  generator} uses this to generate a CPOG file.
The {\em proof checker} verifies the equivalence of the CNF and CPOG representations.
The {\em ring evaluator} computes
a standard or weighted model count from the POG\@ representation.
As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
moves the root of trust away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will be caught by the proof checker.

@@

We evaluate our toolchain using benchmark
formulas from the 2022 standard and weighted model competitions.  Our
tools handle all but the largest decision-DNNF graphs generated
by \dfour{}.  We also measure the benefits of several optimizations as
well as the relative performance of the verified checker with one
designed for high performance and capacity.

@@

Generating proofs of unsatisfiability in SAT solvers has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently checked
\cite{RAT,wetzler14_drattrim}.
A number of formally verified checkers have been developed within different verification frameworks~\cite{cruz-cade-2017,lrat,Lammich:20,Tan:2021}.
The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
Our work builds on the well-established technology and tools associated with clausal proof systems,
but we require features not found in proofs of unsatisfiability.
Our proofs construct a new propositional formula, and we must verify
%both
that it
%satisfies a set of rules and that it
is equivalent to the input formula.  This requires verifying
additional proof steps, including clause deletion steps, and subtle
invariants, as described in Sections~\ref{sect:cpog} and
\ref{section:formally-verified-toolchain}.

Capelli et al.\ developed a version of the \dfour{} knowledge compiler
that generates a certificate in a proof system that is itself based on
decision-DNNF~\cite{capelli:sat:2019,capelli:aaai:2021}.  Their
modified version of \dfour{} generates additional annotations to the
compiled representation to enable checking for equivalence with the
input formula.  They must also track cases where the knowledge
compiler detects that it can reuse a previously generated subgraph,
yielding a DAG rather than a tree representation.  Their proof checker
then must perform a number of checks that we incorporate into our
proof generator.  Our framework has the advantage of being independent
from the means by which the proofs are generated with proof checking
based on very elementary clausal reasoning steps.

Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithms commonly used by model counters.  They present
a series of lemmas and a theorem stating that their checking framework
is sound.  They also present experimental results for cases where they
modify existing model counters to generate proof traces, and where
they generate proof traces from decision-DNNF graphs.
%% They have not formally verified their
%%proof framework or created a formally verified checker.
Our toolchain is not directly comparable to theirs---they directly
certify the numeric value produced by a model counter, while we
certify that the graph generated by a knowledge compiler is logically
equivalent to the input formula.  We can then validate a standard or
weighted count from that representation.
%% Their tool
%% chain can be applied to some programs that are pure model counters,
%%while ours can be applied to pure knowledge compilers.
Nonetheless,
we compare the performance of our toolchain to theirs as part of our
experimental evaluation in Appendix~\ref{app:experiments}.

Both the

@@

POGs generalize decision-DNNF graphs in two ways:
\begin{itemize}
\item They allow negation of arbitrary nodes in the graph, not just
  variables.  
  %, and so we allow it in the interest of generality.
%%  Negation could be useful in some contexts, such as when
%%  converting binary decision diagrams (BDDs) having complemented
%%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.
\item As with d-DNNF~\cite{darwiche:aaai:2002,darwiche:jair:2002}, they allow arbitrary arguments to a $\por$ operation, as long as
  these have disjoint models.  By contrast, each
  sum node $\nodeu$ in a decision-DNNF must have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
%%Our generalization is provided to allow an
%%  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
%%  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
The generalizations encompassed by POGs have also been referred to as {\em deterministic decomposable circuits} (d-Ds)~\cite{monet:amw:2018}.
Our current proof generator only works for knowledge compilers
generating decision-DNNF representations, but these generalizations
allow for future extensions, while maintaining the ability to
efficiently perform ring evaluation.

@@

%% The proof system is based on extended
%% resolution~\cite{Tseitin:1983} with reverse unit propagation
%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
%% proving that a set of clauses logically implies another clause.
%% Alternate version
%% knowledge compilation.  The proof system is based on extended
%% resolution~\cite{Tseitin:1983} with reverse unit propagation
%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
%% proving that a set of clauses logically implies another clause.
%%
%% As an example, consider the formula $\varphi = (\obar{a} \lor b) \land (\obar{b} \lor c \lor \obar{e}) \land (\obar c \lor d)$
%% and a RUP step to derive the target clause $(\obar{a} \lor d \lor \obar{e})$ from the three clauses from $\varphi$.
%% A RUP proof would take the following form.  We start with the assignment that falsifies the target clause. This assignment
%% is extended by clauses in $\varphi$ that become unit and eventually the empty clause. In the illustration below, the
%% involved clauses (antecedents) are shown in the order that they became unit / falsified.
%%
%% \begin{center}
%%   \begin{tabular}{lcccc}
%%          & \makebox[15mm]{Target}    & \makebox[10mm]{} & \makebox[10mm]{Antecedents} & \makebox[10mm]{} \\
%%   Clause & $\obar{a} \lor d \lor \obar{e}$ & $\obar{c} \lor d$  & $\obar{e} \lor \obar{b} \lor c$   & $\obar{a} \lor b$ \\
%%   \midrule
%%   Units  & $a$, $\obar{d}$, $e$      &  $\obar{c}$             & $\obar{b}$                         & $\nil$ \\
%%   \end{tabular}
%% \end{center}
%%
%% RUP is an alternative formulation of resolution.  For target clause
%% $C$, it can be seen that applying resolution operations to the
%% antecedent clauses from right to left will derive a clause $C'$ such
%% that $C' \subseteq C$.  By {\em subsumption} %~\cite{philipp:lpar:2017},
%% we then have $C' \rightarrow C$.  Compared to listing each resolution
%% operation as a separate step, using RUP as the basic proof step makes
%% the proofs more compact.
%%
%% As is shown in Figure~\ref{fig:chain}
%% two programs are involved in producing a certified compilation of a formula:
%% \begin{itemize}
%% \item The {\em proof generator} produces a CPOG file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
%% As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a decision-DNNF representation of the formula.
%% \item The {\em proof checker} ensures that all of the proof conditions are satisfied.
%% \end{itemize}
%% >>>>>>> 9c1793da858d00e5d4022d6d33c009848ba2083b

@@

The sequence of operator declarations, asserted clauses, and
clause deletions represents a systematic transformation of the input formula
into a POG formula.  Validating all of these steps serves to prove that
POG $P$ is logically equivalent to the input formula.
At the completion of the proof, the following \textsc{final conditions} must hold:
\begin{enumerate}
\item There is exactly one remaining clause that was added via RUP
  addition, and this is a unit clause consisting of root literal $r$.
\item All of the input clauses have been deleted.
\end{enumerate}
In other words, at the end of the proof it must hold that the active clauses be exactly those
in $\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting
of unit clause $\{r\}$ and the defining clauses for the nodes, providing a Tseitin encoding of $P$. Recognizing that
any total assignment to the input variables implicitly defines the assignments to the extension variables,
we can see that $\pogformula$ is the clausal encoding of $\phi_\noder$.
Let $\inputformula$ denote the input formula.
The sequence of clause addition steps provides a {\em forward implication} proof that
$\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$.  That is, any total
assignment $\assign$ satisfying the input formula must also satisfy
the formula represented by the POG\@.
Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must
falsify $\phi_\noder$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
that
$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.

@@

\section{Generating CPOG from decision-DNNF}
\label{section:generating:cpog}

%%Figure~\ref{fig:chain} illustrates one method of generating a CPOG
%%file from an input formula in conjunctive normal form.  Our toolchain
%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
%%but it would also work for other tools that use a top-down search
%%strategy to generate a decision-DNNF representation.
A decision-DNNF
graph can be directly translated  into a POG,
although in doing this conversion,
our program also performs simplifications to
eliminate Boolean constants.
Except in degenerate cases,
%where the formula is unsatisfiable or a tautology,
we can therefore assume
that the POG does not contain any constant nodes.
In addition, negation is only
applied to variables, and so we can merge the negation operations into the terminal nodes,
viewing the POG as consisting
of {\em literal} nodes corresponding to input variables and their negations, along with
{\em nonterminal} nodes, which can be further classified as {\em product} and {\em sum} nodes.

@@

The process for generating such a proof depends on the form of node $\nodeu$:
\begin{enumerate}
\item If $\nodeu$ is a literal $\lit'$, then the formula
  $\simplify{\psi}{\passign}$ must consist of the single unit clause
  $C = \{\lit'\}$, such that any $C' \in \prov_{\passign}(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
  Any of these can
  serve as the target clause.
\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
  then, since the node originated from a decision-DNNF graph, there must be
  some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
  product node containing a literal node for $x$ as a child.  In either case, we
  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
  This will cause the addition of the target clause
  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
Similarly, either $\nodeu_0$ is a literal node for $\obar{x}$ or $\nodeu_0$ is a product node containing a literal node for $\obar{x}$ as
  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
  causing the addition of the target clause
  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
  These recursive results can be combined with the second and third defining clauses for $\nodeu$
(see Table~\ref{tab:defining}(B))
  to generate the target clause for $\nodeu$.
\item If $\nodeu$ is a product node, then we can divide its children
  into a set of literal nodes $\lambda$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
  \begin{enumerate}
    \item For each literal
  $\lit \in \lambda$, we must prove that any total assignment $\alpha$, such that
  $\passign \subset \alpha$ has $\alpha(\lit) = 1$.  In some
  cases, this can be done by simple Boolean constraint propagation (BCP).
  In other cases, we must prove that the formula
  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
  do so by writing the formula to a file, invoking a proof-generating
  SAT solver, and then converting the generated unsatisfiability proof
  into a sequence of clause additions in the CPOG file.
  (The solver is constrained to only use RUP inference rules, preventing it from introducing extension variables.)
\item For a single nonterminal child ($k = 1$), we recursively call
  $\validate \left(\nodeu_1, \passign \cup \lambda, \psi\right)$.
\item For multiple nonterminal children ($k > 1$),
  it must be the case that the clauses in
  $\gamma = \simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
  literals.
  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov_{\passign}(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
  We recursively call
  $\validate \left(\nodeu_i, \passign \cup \lambda, \psi_i\right)$.
\end{enumerate}
  We then generate the target clause for node $\nodeu$,
creating the hint by combining the results from the BCP and SAT calls for
  the literals, the recursively computed target clauses, and all but
  the first defining clause for node $\nodeu$
(see Table~\ref{tab:defining}(A)).
\end{enumerate}
Most of these steps involve a polynomial number of
operations per recursive call, with the exception of those that call
a SAT solver to validate a literal.

@@

Each input clause deletion can be done as a single RUP step, based
on an algorithm to test for clausal entailment in d-DNNF graphs~\cite{capelli:sat:2019,darwiche:jair:2002}.  The
proof generator constructs the hint sequence from the defining
clauses of the POG nodes via a single, bottom-up pass through the
graph.  The RUP deletion proof for input clause $C$ effectively proves that any
total assignment $\assign$ that satisfies the POG definition clauses
but does not satisfy $C$ will yield
$\assign(r) = 0$.  It starts with the set of literals
$\{ \obar{\lit} \mid \lit \in C\}$, describing the required condition for
assignment $\assign$ to falsify clause $C$.
It then
adds literals via unit propagation until a
conflict arises.    Unit literal $r$ gets
added right away, setting up a potential conflict.

@@

We summarize the results of our experiments here.  
Appendix~\ref{app:experiments} provides a more complete description. For our
evaluation, we used the public benchmark problems from the 2022
standard and weighted model competitions.  We found that there were
180 unique CNF files among these, ranging in size from 250 to
2,753,207 clauses.
%%We ran our programs on a processor with 64~GB of
%%memory and having an attached high-speed, solid-state
%%disk.
With a runtime limit of 4000 seconds, \dfour{} completed for 124 of the
benchmark problems.  Our proof generator was able to convert all of
these into POGs, with their declarations ranging from 304 to
2,761,457,765 (median 848,784) defining clauses.
%(The
%maximum count overflowed the overflowed the 32-bit signed integers we
%used to represent clause identifiers in both the proof generator and
%checker.)
We ran our proof generator with a time limit of 10,000 seconds.
It was able to generate full proofs for 108 of the problems and
one-sided proofs for an additional 9 of them, leaving just 7 with no
verification.  The prototype checker successfully verified all of the generated
proofs.  The longest runtime for the combination of proof generator
and checker for a full proof was 13,145 seconds.
It is worth noting that, to date, we have not found
any errors in the decision-DNNF graphs generated by \dfour{}.

@@

This paper demonstrates a method for certifying the equivalence of two
different representations of a Boolean formula: an input formula
represented in conjunctive normal form, and a compiled representation
that can then be used to extract useful information about the formula,
including its weighted and unweighted model counts.  It builds on the
extensive techniques that have been developed for clausal
proof systems, including extended resolution and reverse unit propagation, as well as established tools, such as
proof-generating SAT solvers and \dtrim{}.
